---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(tibble.width = Inf)  
```

# eyemovements

<!-- badges: start -->
<!-- badges: end -->

Tools for processing raw eye-tracking samples and detecting fixations and saccades.


## Installation

You can install the package from Github by running the following code:

```{r installation,  eval=FALSE}
# Check if "devtools" is installed:
if('devtools' %in% rownames(installed.packages())==FALSE){
    install.packages('devtools')
    library(devtools)
  }else{
    library(devtools)
  }
devtools::install_github("martin-vasilev/eyemovements")

```

## Sample data:

The package has some in-built datasets to use as an example:

1. `"Oz" dataset`, containing the raw samples of a participant reading a page from the "Little Wizard Stories of Oz" (Slattery & Vasilev, 2019). Data was recorded with an Eyelink 1000+ at 1000 Hz (monocular, from the right eye):

```{r Oz_data}
library(eyemovements)
dat= data_Oz
str(dat)
```

When using your own data, note that it should contain the same columns (`time`, `x`, and `y`). At the moment, binocular data is not supported, but as a temporary solution, you can average out the x and y positions of the two eyes, take just the right eye, or run the data separately for each eye.

## Event detection using the Identification by Velocity Threshold (I-VT) method

Some saccade detection algorithms such as I-VT require the use of degrees of visual angle (dva). For these to work, the degrees need to be calculated for your design. For this you need to know:

- The physical monitor size in cm, provided as c(x, y) vector
- Screen resolution in pixels, provided as c(x, y) vector
- Viewing distance (in cm)- distance between the participant's eye and the monitor

These can then be used in the `VisualAngle()` function to calculate the dva per pixel:

```{r Visual angle}

deg<- VisualAngle(screen_size_cm = c(x = 53, y = 30),
  resolution_px = c(x = 1920, y = 1080),
  viewing_dist_cm = 82)
deg

```

Using the dva values above, we can parse the data for fixations/ saccades using the `IVT()` function. The function outputs fixations by default (`return_saccade= FALSE`). However, saccade data can be alternatively outputted by setting `return_saccades= TRUE`. The sensitivity of the saccade detection algorithm is determined by the   `vel_threshold` argument (in degrees per second). By default, this is set to 30 deg/s (same default that Eyelink uses for "cognitive research"). If working with more oculomotor tasks (e.g., pro-saccade/anti-saccade), this can be set a bit higher (e.g., 35-40 deg/s). However, users can change it to any value. Please note that some data smoothing is recommended before parsing the data to improve performance of the algorithm. Sometimes the algorithm can detect very short fixations, mostly due to noise. Users can specify the minimum fixation duration to be detected (`min_fix_dur`). By default, this is set to 50 ms (i.e., all fixations 0-49 ms are discarded). However, users can turn it off if they prefer (`min_fix_dur= NULL`). 

To detect fixations, we can put the desired arguments together with the Oz data:

```{r IVT_fixations}
fix<- IVT(data = dat,
          dva_x = 0.01865554,
          dva_y = 0.01919689,
          vel_threshold = 30,
          min_fix_dur = 50)

head(fix)
```
The function outputs the following variables for fixations:

- `fix_id`: fixation ID in the detected sequence
- `fix_start`: start of fixation timestamp
- `fix_end`: end of fixation timestamp
- `x`: mean x pixel coordinate of the samples making up the fixation
- `y`: mean y pixel coordinate of the samples making up the fixation
- `fix_dur`: fixation duration (defined as `fix_end`- `fix_start`)
- `avg_velocity`: average velocity (in deg/s) of the samples making up the fixation
- `n_samples`: number of samples making up the fixation

To detect saccades, we can use the same arguments but set `return_saccades` to `TRUE`. Note that the function still performs filtering of very short saccades that are likely noise. This is done with the `min_sacc_amplitude` argument. By default, saccades that are shorter than 0.15 deg in amplitude are discarded (same default that the Eyelink parser uses). However, this can be turned off by setting `min_sacc_amplitude` to `NULL` or `0`.

```{r IVT_saccades}

sacc<- IVT(data = dat,
          dva_x = 0.01865554,
          dva_y = 0.01919689,
          vel_threshold = 30, 
          return_saccades = TRUE)

head(sacc)
```

The function outputs the following variables for fixations:

- `sacc_id`: saccade ID in the detected sequence
- `sacc_start`: start of saccade timestamp
- `sacc_end`: end of saccade timestamp
- `sacc_dur`: saccade duration (defined as `sacc_end`- `sacc_start`)
- `x_start`: horizontal gaze position at saccade onset, in pixels.
- `y_start`: vertical gaze position at saccade onset, in pixels
- `x_end`: horizontal gaze position at saccade offset, in pixels
- `y_end`: vertical gaze position at saccade offset, in pixels
- `sacc_amplitude_deg`: saccade amplitude in degrees of visual angle, computed from the horizontal and vertical displacement between saccade onset and offset
- `avg_velocity_deg`: mean sample-to-sample velocity during the saccade, in deg/s
- `peak_velocity_deg`: maximum sample-to-sample velocity during the saccade, in degrees/second
- `n_samples`: number of samples contributing to the saccade




