---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# eyemovements

<!-- badges: start -->
<!-- badges: end -->

Tools for processing raw eye-tracking samples and detecting fixations and saccades.


## Installation

You can install the package from Github by running the following code:

```{r installation,  eval=FALSE}
# Check if "devtools" is installed:
if('devtools' %in% rownames(installed.packages())==FALSE){
    install.packages('devtools')
    library(devtools)
  }else{
    library(devtools)
  }
devtools::install_github("martin-vasilev/eyemovements")

```

## Sample data:

The package has some in-built datasets to use as an example:

1. `"Oz" dataset`, containing the raw samples of a participant reading a page from the "Little Wizard Stories of Oz" (Slattery & Vasilev, 2019). Data was recorded with an Eyelink 1000+ at 1000 Hz (monocular, from the right eye):

```{r example}
library(eyemovements)
dat= data_Oz
str(dat)
```

