---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(tibble.width = Inf)  

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

## load libraries:
library(tidyverse)
library(patchwork)
```

# eyemovements

<!-- badges: start -->
<!-- badges: end -->

Tools for processing raw eye-tracking samples and detecting fixations and saccades.


## Installation

You can install the package from Github by running the following code:

```{r installation,  eval=FALSE}
# Check if "devtools" is installed:
if('devtools' %in% rownames(installed.packages())==FALSE){
    install.packages('devtools')
    library(devtools)
  }else{
    library(devtools)
  }
devtools::install_github("martin-vasilev/eyemovements")

```

## Sample data:

The package has some in-built datasets to use as an example:

1. `"Oz" dataset`, containing the raw samples of a participant reading a page from the "Little Wizard Stories of Oz" (Slattery & Vasilev, 2019). Data was recorded with an Eyelink 1000+ at 1000 Hz (monocular, from the right eye):

```{r Oz_data}
library(eyemovements)
dat= data_Oz
str(dat)
```

When using your own data, note that it should contain the same columns (`time`, `x`, and `y`), **using the exact same spelling**. Binocular data is not yet supported, but as a temporary solution, you can average out the x and y positions of the two eyes, take just the right eye, or run the data separately for each eye.

## Data smoothing

Before detecting events (saccades, fixations) it is usually helpful to filter the raw gaze positions to reduce noise. The `SmoothSamples()` function provides several algorithms for data smoothing such as Mean/Median moving window, Savitzky-Golay, etc. The exact filter can be specified with the `method` argument in the function (e.g., `Mean`, `Median`, `SG`).

For mean/median filtering, you can provide the moving window size. For data with high temporal resolution (e.g., 500-1000Hz), 5-sample moving window tends to work well. This can be reduced to 3 for low-sampling applications (though note that the value has to be odd).

```{r smoothing1}
# Moving average filtering:
filtered_dat_mean<- SmoothSamples(data = data_Oz, method = "Mean", window_size = 5)

# Moving median filtering:
filtered_dat_median<- SmoothSamples(data = data_Oz, method = "Median", window_size = 5)

```

Other filters such as Savitzky-Golay can also be used. Here, the filter order (`p`) and length (`n`) need to be specified. 


```{r smoothing2}
filtered_dat_SG<- SmoothSamples(data = data_Oz, method = "SG", p= 5, n= 23)
```

```{r smooth_plot, echo=F}

f1<- filtered_dat_mean[1:250,]
f2<- filtered_dat_median[1:250,]
f3<- filtered_dat_SG[1:250,]
gt<- data_Oz[1:250,]

f1$filter<- "Mean"
f2$filter<- "Median"
f3$filter<- "Savitzky-Golay"
gt$filter<- "Raw samples"

f1$time<- 1:nrow(f1)
f2$time<- 1:nrow(f2)
f3$time<- 1:nrow(f3)
gt$time<- 1:nrow(gt)

f<- rbind(f1, f2, f3, gt)

x<- f%>%
  mutate(filter= fct_relevel(filter, 'Raw samples',
                             'Mean', 'Median', 'Savitzky-Golay'))%>%
  ggplot(aes(x = time)) +
#  geom_point(aes(y = x), colour = "grey70", size = 0.1, alpha = 0.5) +
  geom_line(aes(y = x, colour = filter), linewidth = .6) +
  theme_bw()+
  theme(legend.position = 'none')

y<- f%>%
  mutate(filter= fct_relevel(filter, 'Raw samples',
                             'Mean', 'Median', 'Savitzky-Golay'))%>%
  ggplot(aes(x = time)) +
#  geom_point(aes(y = x), colour = "grey70", size = 0.1, alpha = 0.5) +
  geom_line(aes(y = y, colour = filter), linewidth = .6) +
  theme_bw()+
  theme(legend.position = 'right')

```

```{r, out.width="70%", echo=FALSE}
x/y
```


## Event detection (Saccades/ Fixations)

Below we outline the available methods in the package for detecting fixations and saccades.

### Identification by Velocity Threshold (I-VT) method

One of the most common ways of identifying saccades/ fixations is based on their velocities. This relies on the assumption that sample-to-sample velocities are low during fixations but high during saccades. Velocity-based algorithms typically use a velocity threshold (which can be either fixed or adaptive). Samples that exceed the threshold are deemed to be in a saccade, whereas samples that are under the threshold are deemed to be in a fixation. The I-VT method uses a fixed velocity threshold, which can be specified by the user.

Some saccade detection algorithms such as I-VT require the use of degrees of visual angle (dva). For these to work, the degrees need to be calculated for your design. For this you need to know:

- The physical monitor size in cm, provided as c(x, y) vector
- Screen resolution in pixels, provided as c(x, y) vector
- Viewing distance (in cm)- distance between the participant's eye and the monitor

These can then be used in the `VisualAngle()` function to calculate the dva per pixel:

```{r Visual angle}

deg<- VisualAngle(screen_size_cm = c(x = 53, y = 30),
  resolution_px = c(x = 1920, y = 1080),
  viewing_dist_cm = 82)
deg

```

Using the dva values above, we can parse the data for fixations/ saccades using the `IVT()` function. The function outputs fixations by default (`return_saccade= FALSE`). However, saccade data can be alternatively outputted by setting `return_saccades= TRUE`. The sensitivity of the saccade detection algorithm is determined by the   `vel_threshold` argument (in degrees per second). By default, this is set to 30 deg/s (same default that Eyelink uses for "cognitive research"). If working with more oculomotor tasks (e.g., pro-saccade/anti-saccade), this can be set a bit higher (e.g., 35-40 deg/s). However, users can change it to any value. Please note that some data smoothing is recommended before parsing the data to improve performance of the algorithm. Sometimes the algorithm can detect very short fixations, mostly due to noise. Users can specify the minimum fixation duration to be detected (`min_fix_dur`). By default, this is set to 50 ms (i.e., all fixations 0-49 ms are discarded). However, users can turn it off if they prefer (`min_fix_dur= NULL`). 

To detect fixations, we can put the desired arguments together with the Oz data:

```{r IVT_fixations}
fix_IVT<- IVT(data = filtered_dat_mean,
          dva_x = 0.01865554,
          dva_y = 0.01919689,
          vel_threshold = 30,
          min_fix_dur = 50)

head(fix_IVT)
```
The function outputs the following variables for fixations:

- `fix_id`: fixation ID in the detected sequence
- `fix_start`: start of fixation timestamp
- `fix_end`: end of fixation timestamp
- `x`: mean x pixel coordinate of the samples making up the fixation
- `y`: mean y pixel coordinate of the samples making up the fixation
- `fix_dur`: fixation duration (defined as `fix_end`- `fix_start`)
- `avg_velocity`: average velocity (in deg/s) of the samples making up the fixation
- `n_samples`: number of samples making up the fixation

To detect saccades, we can use the same arguments but set `return_saccades` to `TRUE`. Note that the function still performs filtering of very short saccades that are likely noise. This is done with the `min_sacc_amplitude` argument. By default, saccades that are shorter than 0.15 deg in amplitude are discarded (same default that the Eyelink parser uses). However, this can be turned off by setting `min_sacc_amplitude` to `NULL` or `0`.

```{r IVT_saccades}

sacc_IVT<- IVT(data = filtered_dat_mean,
          dva_x = 0.01865554,
          dva_y = 0.01919689,
          vel_threshold = 30, 
          return_saccades = TRUE)

head(sacc_IVT)
```

The function outputs the following variables for fixations:

- `sacc_id`: saccade ID in the detected sequence
- `sacc_start`: start of saccade timestamp
- `sacc_end`: end of saccade timestamp
- `sacc_dur`: saccade duration (defined as `sacc_end`- `sacc_start`)
- `x_start`: horizontal gaze position at saccade onset, in pixels.
- `y_start`: vertical gaze position at saccade onset, in pixels
- `x_end`: horizontal gaze position at saccade offset, in pixels
- `y_end`: vertical gaze position at saccade offset, in pixels
- `sacc_amplitude_deg`: saccade amplitude in degrees of visual angle, computed from the horizontal and vertical displacement between saccade onset and offset
- `avg_velocity_deg`: mean sample-to-sample velocity during the saccade, in deg/s
- `peak_velocity_deg`: maximum sample-to-sample velocity during the saccade, in degrees/second
- `n_samples`: number of samples contributing to the saccade

### Identification by Dispersion Threshold (I-DT) method

An alternative to velocity-based methods is to use sample-to-sample dispersion (i.e., how widely dispersed individual samples are in space). This method relies on the assumption that dispersion will be low during fixations because the eyes will oscillate around the same narrow area and samples will be clustered closely in space. This can be easily by plotting the raw x and y samples. The "blobs" of samples (where a lot of them occur in the same small area) are the fixations.

```{r, blob_plot, out.width="50%", echo=FALSE}
filtered_dat_mean[1:5000,]%>%
  ggplot(aes(x = x, y = y))+
  geom_point(size= 0.15)+
  theme_bw(16)
```

The I-DT method works by starting with a minimum fixation window (`window_threshold`), set in the package by default to 50ms for compatibility with I-VT. The algorithm checks if the samples within the fixation window exceed a fixed dispersion threshold (`disp_threshold`, set by default to 1 deg). If this is not the case, then the window keeps "growing" to the right by adding additional samples until the threshold is exceeded. The fixation is then terminated and recorded. If the initial window already exceeds the threshold, then the first sample is removed on an iterative basis until the dispersion falls below the threshold. The end point of the fixation is then determined by the same process, as described above.


To detect fixations, we can use the `IDT()` function with the mean-filter Oz data:

```{r IDT_fixations}
fix_IDT<- IDT(data = filtered_dat_mean,
          dva_x = 0.01865554,
          dva_y = 0.01919689,
          window_threshold = 50,
          disp_threshold = 1)

head(fix_IDT)
```

The function outputs the following variables for fixations:

- `fix_id`: fixation ID in the detected sequence
- `fix_start`: start of fixation timestamp
- `fix_end`: end of fixation timestamp
- `x`: mean x pixel coordinate of the samples making up the fixation
- `y`: mean y pixel coordinate of the samples making up the fixation
- `fix_dur`: fixation duration (defined as `fix_end`- `fix_start`)
- `fix_dispersion_deg`: average dispersion (in deg) of the samples making up the fixation
- `n_samples`: number of samples making up the fixation

The `IDT()` function can also return saccades, which are defined as intervals of non-fixations. To do so, `return_saccades` needs to be set to `TRUE`. This is provided mostly for comparability purposes. However, some caution needs to be exercised, as the I-DT method is not optimised for saccade detection. 

```{r IDT_saccades}

sacc_IDT<- IDT(data = filtered_dat_mean,
          dva_x = 0.01865554,
          dva_y = 0.01919689,
          window_threshold = 50,
          disp_threshold = 1, 
          return_saccades = TRUE)

head(sacc_IDT)
```

The function outputs the same saccade variables as I-VT (see above).


### Identification by Engbert and Kliegl's (2003) method for (micro)saccades

The `EngbertKliegl03()` function implements the velocity-based algorithm described by Engbert and Kliegl (2003). The method was originally developed for detecting microsaccades during fixation, but it can also be useful as an adaptive saccade-detection method because the velocity threshold is estimated from the noise level in the data.

Unlike the standard I-VT method, this algorithm does not require the user to specify a fixed velocity threshold. Instead, horizontal and vertical velocities are estimated using a 5-sample moving window, and saccades are detected as samples whose velocity falls outside an adaptive elliptical threshold. The size of this threshold is controlled by the `lambda` argument. Engbert and Kliegl (2003) used `lambda = 6`, which is also the default here.

As with the other algorithms, the input data must contain the columns `time`, `x`, and `y`, where `time` is in milliseconds and `x`/`y` are gaze positions in pixels.

Although the function can be used without converting pixels to degrees of visual angle, it is recommended to provide `dva_x` and `dva_y` when saccade amplitude and velocity are of interest. If these are supplied, the gaze positions are still returned in pixels, but saccade amplitudes and velocities are returned in degrees and degrees/second.

```{r EngbertKliegl_fixations}
fix_EK03 <- EngbertKliegl03(
  data = filtered_dat_mean,
  dva_x = 0.01865554,
  dva_y = 0.01919689,
  lambda = 6,
  min_fix_dur = 50
)

head(fix_EK03)
```

By default, the function returns fixations. The output contains:

- `fix_id`: fixation ID in the detected sequence
- `fix_start`: start of fixation timestamp
- `fix_end`: end of fixation timestamp
- `fix_dur`: fixation duration, defined as `fix_end - fix_start`
- `x`: mean horizontal gaze position during the fixation, in pixels
- `y`: mean vertical gaze position during the fixation, in pixels
- `avg_velocity`: mean velocity during the fixation
- `n_samples`: number of samples contributing to the fixation

To return saccades instead, set `return_saccades = TRUE`.

```{r EngbertKliegl_saccades}
sacc_EK03 <- EngbertKliegl03(
  data = filtered_dat_mean,
  dva_x = 0.01865554,
  dva_y = 0.01919689,
  lambda = 6,
  return_saccades = TRUE
)

head(sacc_EK03)
```

When `return_saccades = TRUE`, the function returns:

- `sacc_id`: saccade ID in the detected sequence
- `sacc_start`: start of saccade timestamp
- `sacc_end`: end of saccade timestamp
- `sacc_dur`: saccade duration, defined as `sacc_end - sacc_start`
- `x_start`: horizontal gaze position at saccade onset, in pixels
- `y_start`: vertical gaze position at saccade onset, in pixels
- `x_end`: horizontal gaze position at saccade offset, in pixels
- `y_end`: vertical gaze position at saccade offset, in pixels
- `sacc_amplitude_deg`: saccade amplitude in degrees of visual angle
- `avg_velocity_deg`: mean velocity during the saccade, in degrees/second
- `peak_velocity_deg`: peak velocity during the saccade, in degrees/second
- `n_samples`: number of samples contributing to the saccade

Very small saccades can be removed using the `min_sacc_amplitude` argument. By default, this is set to `0.15`, meaning that saccades smaller than 0.15 degrees are discarded when `dva_x` and `dva_y` are supplied. To keep all detected saccades, set:

```{r EngbertKliegl_all_saccades, eval=FALSE}
sacc_EK03 <- EngbertKliegl03(
  data = filtered_dat_mean,
  dva_x = 0.01865554,
  dva_y = 0.01919689,
  lambda = 6,
  return_saccades = TRUE,
  min_sacc_amplitude = 0
)
```

The value of `lambda` controls how conservative the detection is. Smaller values make the algorithm more sensitive and will detect more saccades. Larger values make the algorithm more conservative and will detect fewer saccades.

```{r EngbertKliegl_lambda_compare}
sacc_EK03_lam6 <- EngbertKliegl03(
  data = filtered_dat_mean,
  dva_x = 0.01865554,
  dva_y = 0.01919689,
  lambda = 6,
  return_saccades = TRUE
)

sacc_EK03_lam12 <- EngbertKliegl03(
  data = filtered_dat_mean,
  dva_x = 0.01865554,
  dva_y = 0.01919689,
  lambda = 12,
  return_saccades = TRUE
)

nrow(sacc_EK03_lam6)
nrow(sacc_EK03_lam12)
```

In general, `lambda = 6` is a sensible starting point if the goal is to reproduce Engbert and Kliegl's original microsaccade-detection approach. For noisier data, or for applications where very short events are undesirable, higher values may be more appropriate. For cognitive research (e.g., reading), the algorithm performs better with values around 10-12.



